Dutch, PPs, ʀ-pronouns

### 2019-12-23 paper.tex
maandagmiddag
sectie 2 ook af, alleen nog 3

\begin{table}[ht]
	\center
	\caption {Distribution between \tit{waarmee} and \tit{met wat}}
	\begin{minipage}{0.45\linewidth}
		\begin{tabularx}{\textwidth}{ccc}
		\toprule
                              & \tit{waarmee} & \tit{met wat} \\
		\midrule
    m:\tsc{acc}, e:\tsc{ins}  & ✔             & *             \\
    m:\tsc{ins}, e:\tsc{acc}  & *             & ✔             \\
    \bottomrule
\end{tabularx}
\end{minipage}
\end{table}\label{tbl:distribution}

@@

\ex. Objects of prepositions \label{ex:objprep}
\ag. Ik schilder samen met haar/hem.\\
 I paint together with her/him\\
 `I am painting together with her/him.'\label{ex:prepani}
\bg. *Ik schilder met het.\\
 I paint with a brush\\
 `I am painting with it.'\label{ex:prephet}
\bg. Ik schilder er-mee.\\
 I paint there-with\\
 `I am painting with it.'\label{ex:preper}
\bg. *Ik schilder mee er.\\
 I paint with-there\\
 `I am painting with it.'\label{ex:erprep}

@@

\begin{table}[ht]
	\center
	\caption {Inanimates in Dutch}
	\begin{minipage}{0.45\linewidth}
		\begin{tabularx}{\textwidth}{ccc}
		\toprule
                        & pers. pronouns & wh-pronouns \\
  	\midrule
objects of verbs        & ət             & wat         \\
objects of prepositions & er             & waar        \\
\bottomrule
\end{tabularx}
\end{minipage}
\end{table}\label{tbl:inanimates}

@@

\ex.
\ag. Ik schilder er-mee.\\
 I paint there-with\\
 `I am painting with it.'\label{ex:const1}
\bg. Waar schilder jij mee?\\
 Where paint you with\\
 `What are you painting with?'\label{ex:const2}
\bg. Ik schilder met de kwast [waar jij ook mee schildert].\\
 I paint with the brush {where with} you also paint\\
 `I am painting with the brush that you are painting with too.'\label{ex:const3}
\bg. Ik schilder [waar jij ook mee schildert].\\
 I paint where you also with paint\\
 `I am painting with what you are painting with too.'\label{ex:const4}

In each of these examples the instrumental PP forms a constituent at a certain point in the derivation. In \ref{ex:const1}, the PP forms a proper constituent in the surface order, as shown in \ref{ex:const1stage}. In \ref{ex:const2}, the PP forms a proper constituent before wh- and V2- movement, shown in \ref{ex:const2stage}.
The structure in \ref{ex:const3stage} represents a stage in the derivation of the embedded clauses in \ref{ex:const3} and \ref{ex:const4}.
\footnote{I assume that the antecedent in a free relative is a phonologically empty element, in line with cf. \cite{bresnan1978a,groos1981,himmelreich2017}.}
Again, in the stage, which comes before relative movement of the pronoun to the left periphery of the relative clause, the PP forms a proper constituent.

@@

\ref{ex:grammaticalwat} is not the only construction in which the string \tit{met wat} `with what' appears. I give examples of two more occurrences in \ref{ex:moremetwat}. In \ref{ex:watindef}, \tit{wat} `what' is the \tit{wat} `what' in the so-called \tit{wat voor} `what for'-construction \citep[cf.][]{corver1991}.
In \ref{ex:watindef}, \tit{wat} appears as a quantifier, and it means `some'. In both construction \tit{wat} `what' takes a complement and \tit{met wat} `with what' do not form a proper constituent. The brackets within the examples indicate the constituency.

@@

insert the zero thingy there

I repeat the example in which \tit{schildert} is in the embedded clause.

\tit{Schildert} wants the instrumental PP, so the embedded clause looks like this. Here \tit{waar} and \tit{mee} are both part of the embedded clause. At a certain point in the derivation they formed a proper constituent, i.e. they formed a constituent without any other elements.

Then in the main clause \tit{gekocht} is merged, which selects for an accusative object. Somehow \tit{waarmee} can satisfy that requirement. \tit{Schildert} wants the instrumental PP, so the embedded clause looks like this:

\ex. [waar jij mee schildert]

Here \tit{wat} is part of the embedded clause but \tit{met} is not. It belongs to the main clause. An embedded clause is finished before the main clause is built. At no point in the derivation can \tit{met} \tit{wat} form a proper constituent.

In conclusion:

\ex.
\a. *[[met] [wat]] → [waarmee]
\b. [met [wat X]]
\b. [met [wat [X]]]

How do I connect constituency and different spellouts of the instrumental? Nano: only constituents can spell out.

\section{Connecting constituency and spellout}

Intro: why *\tit{met wat}? Not a filter, not a different paradigm.

\subsection{Korean negation and causative}

\subsection{nanostory of that}

\subsection{waarmee = parallel}

@@

@@

Now let's look at it like this: \tit{waarmee} realizes particular features. Say \tit{waar} expresses A and B and \tit{mee} expresses C and D. Say, A B C and D. \tit{met wat} realizes the same features. Saying that \tit{wat} also realizes A and B and \tit{met} also C and D misses the point of having different realizations. So instead, say \tit{wat} only realizes A and \tit{met} realizes B, C and D. Or \tit{wat} realizes A, B and C and \tit{met} only realizes D. In both cases all features are expresses, but they are distributed differently, allowing for different realizations.\footnote{I did not address the ordering difference of the elements. Yes, this can lead to a difference in meaning.

\ex.
\ag. Ik klim in de boom.\\
 I climb in the tree\\
 `I am climbing in the tree.'
\bg. Ik klim de boom in\\
 I climb the tree in\\
 `I am climbing into the tree.'

Here, however, I will argue that the movement is driven by spellout and is meaningless.}

\begin{table}[ht]
	\center
	\caption {Distribution options}
	\begin{minipage}{0.27\linewidth}
		\begin{tabularx}{\textwidth}{cccc}
		\toprule
    \tsc{a}   & \tsc{b}            & \tsc{c}   & \tsc{d}                           \\
		\midrule
    \multicolumn{2}{c}{\tit{met}}  & \multicolumn{2}{c}{\cellcolor{Gray}\tit{wat}} \\
    \tit{mee} & \multicolumn{3}{c}{\cellcolor{Gray}\tit{waar}}                     \\
    \multicolumn{3}{c}{\tit{mee}}               & \cellcolor{Gray}{\tit{waar}}     \\
    \bottomrule
\end{tabularx}
\end{minipage}
\end{table}

@@

\begin{table}[ht]
	\center
	\caption {\tit{er} < \tit{'t}}
	\begin{minipage}{0.27\linewidth}
		\begin{tabularx}{\textwidth}{cccc}
		\toprule
    \tsc{a}   & \tsc{b}            & \tsc{c}   & \tsc{d}                           \\
		\midrule
    \multicolumn{2}{c}{\tit{met}}  & \multicolumn{2}{c}{\cellcolor{Gray}\tit{wat}} \\
    \multicolumn{3}{c}{\tit{mee}}               & \cellcolor{Gray}{\tit{waar}}     \\
    \bottomrule
\end{tabularx}
\end{minipage}
\end{table}

As a result of \tit{waar} being small, \tit{mee} becomes big. And \tit{mee} spells out more features than \tit{met}.

@@

\begin{itemize}
  \item I still need to incorporate \tit{a} and \tit{w}. They can either come before case or after it.
  \item Right now I do not capture the formal similarity between \tit{met} and \tit{mee}. I have ideas about that:
  \item Some adpositions do not change shape. They should be usable both as prepositions and postpositions. This should be reflected in their lexical entry. A prediction is that there exist adpositions that cannot be used as postpositions. Examples of that are: \tit{zonder}, \tit{midden}, \tit{sinds}, \tit{rond}, \tit{beneden}.
\end{itemize}

\section{Conclusion}

>>>>

\begin{table}[ht]\label{tbl:distribution}
	\center
	\caption {Distribution between \tit{waarmee} and \tit{met wat}}
	\begin{minipage}{0.45\linewidth}
		\begin{tabularx}{\textwidth}{ccc}
		\toprule
                              & \tit{waarmee} & \tit{met wat} \\
		\midrule
    m:\tsc{acc}, e:\tsc{ins}  & ✔             & *             \\
    m:\tsc{ins}, e:\tsc{acc}  & *             & ✔             \\
    \bottomrule
\end{tabularx}
\end{minipage}
\end{table}

@@

\ex. Objects of prepositions \label{ex:objprep}
\ag. Ik schilder samen met haar/hem.\\
 I paint together with her/him\\
 `I am painting together with her/him.'\label{ex:prepani}
\bg. *Ik schilder met 't.\\
 I paint with it\\
 `I am painting with it.'\label{ex:prephet}
\bg. Ik schilder er-mee.\\
 I paint there-with\\
 `I am painting with it.'\label{ex:preper}
\bg. *Ik schilder mee er.\\
 I paint with-there\\
 `I am painting with it.'\label{ex:erprep}

@@

\begin{table}[ht]\label{tbl:inanimates}
	\center
	\caption {Inanimates in Dutch}
	\begin{minipage}{0.7\linewidth}
		\begin{tabularx}{\textwidth}{ccc}
		\toprule
                        & pers. pronouns & wh-pronouns \\
  	\midrule
objects of verbs        & ət             & wat         \\
objects of prepositions & er             & waar        \\
\bottomrule
\end{tabularx}
\end{minipage}
\end{table}

@@

\ex.
\ag. Ik schilder er-mee.\\
 I paint there-with\\
 `I am painting with it.'\label{ex:const1}
\bg. Waar schilder jij mee?\\
 Where paint you with\\
 `What are you painting with?'\label{ex:const2}
\bg. Ik schilder met de kwast [waar jij ook mee schildert].\\
 I paint with the brush where you also with paint\\
 `I am painting with the brush that you are painting with too.'\label{ex:const3}
\bg. Ik schilder [waar jij ook mee schildert].\\
 I paint where you also with paint\\
 `I am painting with what you are painting with too.'\label{ex:const4}

In each of these examples the instrumental PP forms a constituent at a certain point in the derivation. In \ref{ex:const1}, the PP forms a proper constituent in the surface order, as shown in \ref{ex:const1stage}. In \ref{ex:const2}, the PP forms a proper constituent before wh- and V2- movement, shown in \ref{ex:const2stage}.
The structure in \ref{ex:const3stage} represents a stage in the derivation of the embedded clauses in \ref{ex:const3} and \ref{ex:const4}.
\footnote{I assume that the antecedent in a free relative is a phonologically empty element, in line with cf. \cite{bresnan1978a,groos1981,himmelreich2017}. Under that view, \ref{ex:const3} and \ref{ex:const4} are identical, except for that in \ref{ex:const4} modifies phonologically empty element.}
Again, in the stage, which comes before relative movement of the pronoun to the left periphery of the relative clause, the PP forms a proper constituent.

@@

The mismatching free relative in \ref{ex:grammaticalwat} is not the only construction in which the string \tit{met wat} `with what' appears. I give examples of two more occurrences in \ref{ex:moremetwat}. In \ref{ex:watindef}, \tit{wat} `what' is the \tit{wat} `what' in the so-called \tit{wat voor} `what for'-construction \citep[cf.][]{corver1991}.
In \ref{ex:watindef}, \tit{wat} appears as a quantifier, and it means `some'. In both construction \tit{wat} `what' takes a complement and \tit{met wat} `with what' do not form a proper constituent. The brackets within the examples indicate the constituency.

@@

I repeat the mismatching free relative in which \tit{waarmee} `met wat' appears in \ref{ex:mismatchwaarmee}. The predicate \tit{schildert} `paints' combines in the embedded clause with the instrumental PP. The PP forms a proper constituent within the embedded clause, and the PP can be realized as the \tsc{r}-pronoun \tit{waarmee} `with what'.\footnote{I assume that the embedded clause modifies a phonologically empty noun here which is the DP argument of \tit{gekocht} `bought'.}

\exg. Ik heb gekocht [waar jij mee schildert].\\
 I have bought where you with paint\\
 `I bought what you are painting with.'\label{ex:mismatchwaarmee}

Next, we arrive at the mismatching free relative in which \tit{waarmee} `with what' cannot be used, but \tit{met wat} `with what' appears. The embedded clause predicate \tit{gekocht} `bought' combines with an accusative DP. The accusative object of a verb is always \tit{wat} `what', as I showed in \ref{ex:wat}. The instrumental PP only comes into the picture in the main clause, when \tit{schilder} `paint' combines with an instrumental PP. At no point in the derivation does the PP form a proper constituent, and \tit{waarmee} `with what' does not surface.

\exg. Ik schilder met [wat jij hebt gekocht].\\
 I paint with what you have bought\\
 `I paint with what you bought.'\label{ex:mismatchmetwat}

\ref{ex:summaryconst} summarizes what I showed in this section. \tit{Met wat} `with what' can never surface when \tit{met} `with' and \tit{wat} `what' form a proper constituent. It always becomes \tit{waarmee} `with what'. This is schematically shown in \ref{ex:waarmeefr}.
There are other contexts in which \tit{met wat} `with what' appears. This can be either when \tit{wat} `what' takes a complement, or when \tit{wat} is part of the a clause that \tit{met} `with' is not a part of.This last option is schematically showed in \ref{ex:metwatfr}, and it represents the mismatching free relative in \ref{ex:mismatchmetwat}.

\ex.\label{ex:summaryconst}
\a. *[[met] [wat]] → [waarmee]\label{ex:waarmeefr}
\b. [met [wat [X]]]
\b. [met [[wat] [X]]]\label{ex:metwatfr}

This section showed that being a proper constituent is a necessary requirement for \tit{waarmee} `with what' to surface. The next section elaborates on how proper constituency and spellout relate to each other.

\section{\tit{Waarmee} as suppletion}

\tit{Met} `with' is a preposition that combines with full DPs and animate pronouns. \tit{Wat} `what' is a wh-element that appears as subject or object. There is no principled reason why \tit{met wat} `with what' should be ruled out. Previous work on \tsc{r}-pronoun has not given many insights, other than \citet{riemsdijk1978} postulating a filter that prohibits the existence of constituents consisting of prepositions and wh-elements or \citet{koopman2000} who makes reference to a different paradigm.

In this section I start with showing a similar instance in which proper constituency and spellout go hand in hand. Then I show how these facts can be straightforwardly follow from the spellout algorithm in Nanosyntax. The section ends with showing how this logic can be applied to \tit{waarmee} `with what' and \tit{met wat} `with what'.

In what follows I show another instance in which proper constituency influences spellout, suppletive negative forms in Korean as discussed in \citet{chung2007}. The point is that suppletion only goes through when the relevant features form a proper constituent, and not when an additional feature intervenes.

Sentences in Korean can be negated by the two negative prefixes \tit{ani} or \tit{mos}, as shown in \ref{ex:koreannegation}.

\ex.\label{ex:koreannegation}
\ag. ca -n -ta\\
 sleep -\tsc{pres} -\tsc{decl}\\
 `is sleeping'
\bg. mos/an(i) ca -n -ta\\
 \tsc{neg} sleep -\tsc{pres} -\tsc{decl}\\
 `cannot sleep / is not sleeping' \hfill \citep{chung2007}

The verb \tit{al-} `know' does not combine with any of the negative markers, but it it uses the suppletive stem \tit{molu.\tsc{neg}}. This is illustrated in \ref{ex:koreanknow}.

\ex.\label{ex:koreanknow}
\ag. al -n -ta\\
know \tsc{-pres} \tsc{-decl}\\
`know(s)'
\bg. *mos/*an(i) al -n -ta\\
\tsc{neg} know -\tsc{pres} -decl\tsc{}\\
\bg. molu -n -ta\\
\tsc{neg}.know -\tsc{pres} -\tsc{decl}\\
`do(es) not / cannot know'\label{ex:molu} \hfill \citep{chung2007}

When \tit{al-} `know' is causativized before it is negated, the regular stem is used again, shown in \ref{ex:knowcausneg}. \ref{ex:molucaus} illustrates that the suppletive stem \tit{molu-} is no longer grammatical.

\ex.
\ag. al -li-\\
 know -\tsc{caus}\\
 `let know, inform'
\bg. ani/ mos al -li -ess -ta\\
 \tsc{neg} \tsc{neg} know -\tsc{caus} -\tsc{past} -\tsc{decl}\\
 `did not /could not inform'\label{ex:knowcausneg}
\bg. *molu -li -ess -ta\\
 \tsc{neg}.know -\tsc{caus} -\tsc{past} -\tsc{decl}\\\label{ex:molucaus} \hfill \citep{chung2007}

\citet{chung2007} points out that it cannot be phonological adjacency of the negation and \tit{al-} `know' that causes the suppletion in \ref{ex:molu}. The reason for that is that in \ref{ex:knowcausneg} \tit{al-} also follows the negation, but there is no suppletion. \citet{chung2007} suggests crucial difference has to do with in proper constituency.
The structure of \ref{ex:molu} is given in \ref{ex:structuremolu}: Neg and \tit{al-} `know' form a proper constituent. \ref{ex:structurenegmolu} represents the structure of \ref{ex:knowcausneg}. Neg and \tit{al-} `know' no longer form a proper constituent, because \tsc{caus} is also part of the structure as the sister of \tit{al-} `know'.

\ex.
%%%
\a. \begin{forest} boom
[
    [Neg]
    [know]
]
\end{forest}\label{ex:structuremolu}
\b. \begin{forest} boom
[
    [Neg]
    [
        [know]
        [\tsc{caus}]
    ]
]
\end{forest}\label{ex:structurenegmolu}

\citet{caha2009} uses this pattern to argue in favor of a model with phrasal spellout rather than inserting phonology into terminal nodes, specifically Nanosyntax \citep{starke2009}. The idea is that the lexicon contains syntactic structures that are linked to a particular piece of phonology. If Korean has a lexical entry as in \ref{ex:entrymolu}, this can be inserted into the syntactic tree in \ref{ex:structuremolu}.

\ex. \begin{forest} boom
[
    [Neg]
    [know]
]
  {\draw (.east) node[right]{⇔ \tit{molu}}; }
\end{forest}\label{ex:entrymolu}

Important is in the insertion process is the Superset Principle.

\ex. The Superset Principle \citet{starke2009}: \\
A lexically stored tree matches a syntactic node iff the lexically stored tree contains the syntactic node.

This ensures \ref{ex:entrymolu} is not inserted into the structure in \ref{ex:structurenegmolu}, as \tsc{caus} is contained in the syntactic structure but not in the structure in the lexicon in \ref{ex:entrymolu}.

The same logic can be applied to \tit{waarmee} `with what' and \tit{met wat} `with what'. The preposition and wh-element together form a proper constituent, as in \ref{ex:structurewaarmee}, or the embedded CP of the relative clause is a sister of the wh-element, as in \ref{ex:structuremetwat}.

\ex.
\a. \begin{forest} boom
[
    [P]
    [Wh]
]
\end{forest}\label{ex:structurewaarmee}
%%
\b. \begin{forest} boom
[
    [P]
    [
        [Wh]
        [CP
            [...,roof]
        ]
    ]
]
\end{forest}\label{ex:structuremetwat}

Assume we have a lexical entry for \tit{waarmee} `met wat' that look as \ref{ex:entrywaarmee}. This entry can only be inserted into \ref{ex:structurewaarmee}, because the lexical structure matches the syntactic structure. It cannot be into \ref{ex:structuremetwat}, because the lexical structure does not contain the CP.

\ex. \begin{forest} boom
[
    [InsP]
    [WhP]
]
  {\draw (.east) node[right]{⇔ \tit{waarmee}}; }
\end{forest}\label{ex:entrywaarmee}

What I have shown in this section is how proper constituency relates to a change in phonology. However, this analysis missed numerous phonological properties of \tit{waarmee} `with what' and \tit{met wat} `with what', e.g. that they both contain the morpheme \tit{wa}. Under the suppletion analysis above, one can the phonological form of \tit{waarmee} could just as well be completely phonologically unrelated. The point of presenting the problem like this is to illustrate that the distribution between \tit{waarmee} `with what' and \tit{met wat} `with what' can be captured as a matter of spellout.

In the next section I decompose \tit{waarmee} `with what' and \tit{met wat} `with what'. I show that both spell out the same set of features, but the distribution is different.

@@

In this section I investigate the internal structure of \tit{waarmee} `with what' and \tit{met wat} `with what' to capture the phonological similarities between the two forms.

\begin{table}[ht]
	\center
	\caption {\tit{met 't} and \tit{ermee}}
	\begin{minipage}{0.56\linewidth}
		\begin{tabularx}{\textwidth}{ccccccc}
		\toprule
    \tsc{wh}          & \tsc{deix}                         & \tsc{f}4   & \tsc{f}3 & \tsc{f}2  & \tsc{f}1                 & \tsc{thing}                 \\
		\midrule
    \multirow{2}{*}{w}  & \multirow{2}{*}{a}               & \tit{met}  & \multicolumn{4}{c}{\cellcolor{Gray}\tit{(ə)t}}                                \\
                      &                                    & \multicolumn{4}{c}{\tit{mee}}                                & \cellcolor{Gray}\tit{(ə)r}  \\
    \bottomrule
\end{tabularx}
\end{minipage}
\end{table}

The \tsc{r}-pronoun I focus on is the relative pronoun \tit{waarmee} `with what', which is interesting for two reasons. First, just like for all \tsc{r}-pronouns, the wh-element is the locative, but there is no meaning component related to location in `with what'. Second, the preposition \tit{met} `with' changes into \tit{mee} `with when it is combined with an \tsc{r}-pronoun.

What we do not capture is:
formal similarities between waarmee and met wat. I will talk about w, a. What I will not talk about it: mee and met
why we see the locative. I will talk about that.

@@

Now let's look at it like this: \tit{waarmee} realizes particular features. Say \tit{waar} expresses A and B and \tit{mee} expresses C and D. Say, A B C and D. \tit{met wat} realizes the same features. Saying that \tit{wat} also realizes A and B and \tit{met} also C and D misses the point of having different realizations. So instead, say \tit{wat} only realizes A and \tit{met} realizes B, C and D. Or \tit{wat} realizes A, B and C and \tit{met} only realizes D. In both cases all features are expresses, but they are distributed differently, allowing for different realizations.
\footnote{I did not address the ordering difference of the wh-element and the adposition. A topic related to (but not relevant for) this paper is the different positioning of identical adpositions in Dutch. In \ref{ex:dutchin}, \tit{in} changes meaning dependent on whether it procedes or follows the DP, it is respectively locational or directional.

\ex.\label{ex:dutchin}
\ag. Ik klim in de boom.\\
 I climb in the tree\\
 `I am climbing in the tree.'
\bg. Ik klim de boom in.\\
 I climb the tree in\\
 `I am climbing into the tree.'

The different positioning of the adpositions in \tsc{r}-pronouns is driven by spellout and is, therefore, meaningless.}

@@

@@

 still need to incorporate \tit{a} and \tit{w}. They can either come before case or after it.

\section{Conclusion and open issues}

In the analysis laid out above it is a coincidence that \tit{mee} `with' and \tit{met} `with' look so much alike.

Another topic I do not discuss is that in almost all cases the preposition does not change form when it combines with an \tsc{r}-pronoun, e.g. \tit{in}. If this proposal is on the right track, such elements can be used as either a prefix and as a suffix. A lexical entry as in \ref{ex:presuf} would be a candidate for such an element.

\ex. \begin{forest} boom
[XP
    [XP
        [X,roof]
    ]
    [YP
        [Y]
        [\phantom{x}]
    ]
]
{\draw (.east) node[right]{⇔ \tit{in}}; }
\end{forest}\label{ex:presuf}

I leave it to future research to determine whether this is a feasible solution.
